Structural replication of nanoporous media using procedural noise

### 2017-08-17 ms.tex
Fixed spelling of references

The field of synthetic rocks is a large field both scientifically and technologically \cite{biswal2007stochastic,hamzehpour2006development,Reconstructing random media}. 
Methods to generate synthentic rock samples from experimental characterizations may be based on purely statistical or geometrical models, models of physical processes that generate the samples, or combinations of both. 
For small-scale porous media, physical models are often used to generate initial grain packings \cite{pilotti1998generation,scott1969density,herrmann2013physics}, then further mechano-chemical and fluid-mediated processes such as precipitation-dissoluton processes and pressure solution \cite{utter1983pressure}, or fracture and plastic deformation processes are added to include effects of long-term compaction \cite{rutter1983pressure,renard1997pressure}. 
On large scales, reservoir models are populated by statistical models for lithological and geomorphological processes \cite{pyrcz2014geostatistical}, such as river fan formation to generate particular grain size distributions \cite{oltermann1996heterogeneity,lair1994alluvial}. 
More statistical approaches include variations of percolation models \cite{sahimi} or network models \cite{fatt1956network,sahimi1993flow} to characterize and generate uncorrelated or correlated porous media. 
In addition, the high resolution of observational methods allow three-dimensional scans to form the basis for atomic-scale modeling using molecular modeling methods such as molecular dynamics or Monte Carlo methods to study the dynamics of nanoporous media. 
But then procedural or physical models are needed to fill in the atomic scale details, to build many statistical realizations of nanoporous media with atomic-scale details, and to extrapolate to larger scale models.

@@

A much simpler example would be generating a 1 (or 2)-dimensional terrain by choosing random points on a grid and performing a spline interpolation between them to create the illusion of a smooth landscape.
In fact, this is the very essence of procedural noise, which we are focusing on in our model, and the first person to create such a fast algorithm for N-dimensional spline interpolation between pseudo-random points was Ken Perlin.

@@

\subsection{Multiple octaves: creating patterns}
\label{sec:octaves}
The different octaves from procedural noise functions can be combined
to produce a wide range of nature-like patterns. A generic
N-dimensional procedural function $\Phi(\vec x): \mathbb R^N \to
\mathbb R$ could be expressed as
\begin{equation}
\label{eq:procedural}
 \Phi(\vec x) = \Theta \Big(\sum_k^\Omega \kappa (k) P(  f_s (\vec x,k) )) \Big),
 \end{equation}
where $f_s(\vec x,k)$ describes the frequency scale modifier (e.g. $f_s(\vec x,k) =
k\vec x$), $\omega$ is the amount of octaves (modes), $\kappa(k)$ is the octave amplitude (e.g. $\kappa(k) = 1/k$)
and $\Theta(x)$ an overall modifier (e.g. $\Theta(x) = x$ or
$\Theta(x) = \frac{1}{x}$). In a sense, $\kappa(k)$ is similar to Fourier coefficients. 
In its simplest case, summing directly over the octaves with amplitude $1/k$ results in 
\begin{equation}
\label{eq:perlinlinear}
 \Phi(\vec x) = \sum_k^\Omega \frac{1}{k} P( 2k\vec x).
\end{equation}
In one dimension, we show equation \ref{eq:perlinlinear} with 1, 3 and 30 octaves in figure \ref{fig:1dperlin}, while the 2D procedural noise pattern depicted in the lower
part of figure \ref{fig:noise}. Finally, since simplex noise both scales better with higher dimensions and produces less asymmetric effects, we choose to use simplex over regular Perlin noise as the main noise function throughout this paper.

@@

\section{Model}
In this paper, we are interested in developing a model using simplex noise to reproduce nanoporous structures that are as statistically close to physical simulations of porous media.
For a given data set and a model, we wish to estimate the parameters that best represent the data. 
This is performed through a full likelihood analysis in order to obtain the posterior probability distribution of the parameters. 
The model measure of the analysis is the radial distribution function $g(r)$, for which the parameters will be maximised to fit. 
In addition, we also investigate the porosity and surface area of the system. 
The simulated system is confined to a cube of size $\SI{159} {\angstrom}^3$, but since the procedural noise function is continuous on all scales, the box could be of arbitrary size.

@@

Ideally, the model measure should pick up all geometrical properties that are relevant for the physical study of interest. 
For instance, transport in porous media heavily depends on the pore size distribution and porosity, whereas chemical reactons are sensitive to surface area \cite{coussy2011mechanics}.
In this paper, we choose to focus on the radial distribution function $g(r)$ as defined and implemented in LAMMPS \cite{plimpton1995fast}. $g(r)$ describes how density varies as a function of distance from a reference particle, picking up both the intramolecular and intermolecular structure. 
We have chosen to ignore the short range contributions of $g(r)$ since the pore geometry does not depend on the intra-molecular structure of SiO$_2$, but only the density variations at larger distances. 
Hereafter, $g_d(r)$ denotes the RDF calculated from simulated or experimental data, while $g_{M(\bar \theta)}(r)$ represents the RDF based on the procedural noise model $M(\bar \theta)$, where $\bar \theta$ is the set of model parameters.

The method used for calculating porosity of a media is defined as in \cite{gelb1998characterization}, and we obtain the surface area of a porous media by using marching cubes triangulation to sum the surface area of the generated triangles.
As discussed in \cite{gelb1998characterization}, the definition of surface area is ill-defined in atomic simulations, because a clear definition of pore space is needed. 
In the marching cubes triangulation, we create a voxelized grid, a scalar field with values being the distance from the voxel centre to the nearest atom. 
The isovalue threshold defining the surface is chosen to be \SI{2}{\angstrom}, large enough to contribute to a dense pore matrix, but small enough to have most of the available pore space defined as such. 
The important note here is that the same definition of both surface and porosity is being used in all systems that are compared, physical simulations, procedural methods or data.

@@

Note that the surface atoms are not in chemical equilibrium since some neighbours of a surface atom might have been removed. 
In order to use the generated geometry in molecular dynamics, one would need either noble gas atoms which may be stable, or apply passivation step.

>>>>

The field of synthetic rocks is a large field both scientifically and technologically \cite{biswal2007stochastic,hamzehpour2006development,Reconstructing random media}. 
Methods to generate synthentic rock samples from experimental characterizations may be based on purely statistical or geometrical models, models of physical processes that generate the samples, or combinations of both. 
For small-scale porous media, physical models are often used to generate initial grain packings \cite{pilotti1998generation,scott1969density,herrmann2013physics}, then further mechano-chemical and fluid-mediated processes such as precipitation-dissoluton processes and pressure solution \cite{rutter1983pressure}, or fracture and plastic deformation processes are added to include effects of long-term compaction \cite{rutter1983pressure,renard1997pressure}. 
On large scales, reservoir models are populated by statistical models for lithological and geomorphological processes \cite{pyrcz2014geostatistical}, such as river fan formation to generate particular grain size distributions \cite{koltermann1996heterogeneity,blair1994alluvial}. 
More statistical approaches include variations of percolation models \cite{sahimi} or network models \cite{fatt1956network,sahimi1993flow} to characterize and generate uncorrelated or correlated porous media. 
In addition, the high resolution of observational methods allow three-dimensional scans to form the basis for atomic-scale modeling using molecular modeling methods such as molecular dynamics or Monte Carlo methods to study the dynamics of nanoporous media. 
But then procedural or physical models are needed to fill in the atomic scale details, to build many statistical realizations of nanoporous media with atomic-scale details, and to extrapolate to larger scale models.

@@

A much simpler example would be generating a 1 (or 2)-dimensional terrain by choosing random points on a grid and performing a spline interpolation between them to create the illusion of a smooth landscape.
In fact, this is the very essence of procedural noise, which we are focusing on in our model, and the first person to create such a fast algorithm for N-dimensional spline interpolation between pseudo-random points was Ken Perlin\cite{perlin1985image}.

@@

\subsection{Multiple octaves: creating patterns}
\label{sec:octaves}
The different octaves from procedural noise functions can be combined
to produce a wide range of nature-like patterns. A generic
N-dimensional procedural function $\Phi(\vec x): \mathbb R^N \to
\mathbb R$ could be expressed as
\begin{equation}
\label{eq:procedural}
 \Phi(\vec x) = \Theta \Big(\sum_k^\Omega \kappa (k) P(  f_s (\vec x,k) )) \Big),
 \end{equation}
where $f_s(\vec x,k)$ describes the frequency scale modifier (e.g. $f_s(\vec x,k) =
k\vec x$), $\omega$ is the amount of octaves (modes), $\kappa(k)$ is the octave amplitude (e.g. $\kappa(k) = 1/k$)
and $\Theta(x)$ an overall modifier (e.g. $\Theta(x) = x$ or
$\Theta(x) = \frac{1}{x}$). In a sense, $\kappa(k)$ is similar to Fourier coefficients. 
In its simplest case, summing directly over the octaves with amplitude $1/k$ results in 
\begin{equation}
\label{eq:perlinlinear}
 \Phi(\vec x) = \sum_k^\Omega \frac{1}{k} P( 2k\vec x).
\end{equation}
In one dimension, we show equation \ref{eq:perlinlinear} with 1, 3 and 30 octaves in figure \ref{fig:1dperlin}, while the 2D procedural noise pattern is depicted in the lower
part of figure \ref{fig:noise}. Finally, since simplex noise both scales better with higher dimensions and produces less asymmetric effects, we choose to use simplex over regular Perlin noise as the main noise function throughout this paper.

@@

\section{Model}
In this paper, we are interested in developing a model using simplex noise to reproduce nanoporous structures that are statistically similar to physical simulations of porous media.
For a given data set and a model, we wish to estimate the parameters that best represent the data. 
This is performed through a full likelihood analysis in order to obtain the posterior probability distribution of the parameters. 
The model measure of the analysis is the radial distribution function $g(r)$, for which the parameters will be maximised to fit. 
In addition, we also investigate the porosity and surface area of the system. 
The simulated system is confined to a cube of size $\SI{159} {\angstrom}^3$, but since the procedural noise function is continuous on all scales, the box could be of arbitrary size.

@@

Ideally, the model measure should pick up all geometrical properties that are relevant for the physical study of interest. 
For instance, transport in porous media heavily depends on the pore size distribution and porosity, whereas chemical reactons are sensitive to surface area \cite{coussy2011mechanics}.
In this paper, we choose to focus on the radial distribution function (RDF) $g(r)$ as defined and implemented in LAMMPS \cite{plimpton1995fast}. $g(r)$ describes how density varies as a function of distance from a reference particle, picking up both the intramolecular and intermolecular structure. 
We have chosen to ignore the short range contributions of $g(r)$ since the pore geometry does not depend on the intra-molecular structure of SiO$_2$, but only the density variations at larger distances. 
Hereafter, $g_d(r)$ denotes the RDF calculated from simulated or experimental data, while $g_{M(\bar \theta)}(r)$ represents the RDF based on the procedural noise model $M(\bar \theta)$, where $\bar \theta$ is the set of model parameters.

The method used for calculating porosity of a media is defined as in \cite{gelb1998characterization}, and we obtain the surface area of a porous media by using m arching cubes triangulation to sum the surface area of the generated triangles.
As discussed in \cite{gelb1998characterization}, the definition of surface area is ill-defined in atomic simulations, because a clear definition of pore space is needed. 
In the marching cubes triangulation, we create a voxelized grid, a scalar field with values being the distance from the voxel centre to the nearest atom. 
The isovalue threshold defining the surface is chosen to be \SI{2}{\angstrom}, large enough to contribute to a dense pore matrix, but small enough to have most of the available pore space defined as such. 
The important note here is that the same definition of both surface and porosity is being used in all systems that are compared, physical simulations, procedural methods or data.

@@

Note that the surface atoms are not in chemical equilibrium since some neighbours of a surface atom might have been removed. 
In order to use the generated geometry in molecular dynamics, one would need either noble gas atoms which may be stable, or to apply passivation step.
